Text Is All You Need

### 2023-05-04 docs/hes_final/article.qmd
update metrics table imgs

We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc). 
Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
produced comparative results. Each of the models outperformed the other two in a subset of these metrics.

@@

![Model performance metrics (larger values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}

![Model performance metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}

Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability could be a reason to prefer Transformer architecture and resampled dataset.

@@

While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model evaluation. Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.

@@

In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global collection of news publications. We compared the usage of text-only data and data enriched with country labels and social indicators. We experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The results suggest that text embeddings provide a strong enough signal to make such predictions. Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numetic metrics. Transformer-based model generalized better across geographies while statistical model provided a size/cost of deployment advantage.

>>>>

We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results. Each of the models outperformed the other two in a subset of these metrics.

![Model metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}

![Model metrics (large values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}

@@

@@

Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world. While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model evaluation. Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.

@@

In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global collection of news publications. We compared the usage of text-only data and data enriched with country labels and social indicators. We experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The results suggest that text embeddings provide a strong enough signal to make such predictions. Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numeric metrics. The neural network-based models appear to generalize better across geographies while statistical model provided a size/cost of deployment advantage.
